NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2017-12-01 Manuscript/manuscriptPGPE.tex
Updates related to the separation of manuscripts

\title{NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE, PG and PS lipids} %Title of paper

@@

In NMRlipids I and II project we were looking for a MD model
which would correctly reproduce headgroup and glycerol
backbone structures and cation binding for PC lipid bilayers \cite{botan15,catte16}.
Here we extend the same goal for other than PC lipids.
Currently the focus is on PE, PG and PS bilayers and their
mixtures with PC. Experimental data with different amounts of 
added salt is now collected and presented in this manuscript.
Simulations for bilayers containing PG and PS with low salt conditions
are ran with parameters from CHARMM GUI
and compared to experiments. Some order parameters from
simulations seems to be off from experimental values.
Simulation data from other models would be highly useful 
to see if some of the existing models would reproduce the 
experimental order parameters and to analyze different 
conformations predicted by different models respect to 
experiments.

@@

\section{Experimental glycerol backbone and headgroup order parameters for PE, PG and PS lipids}

@@

\section{Glycerol backbone and headgroup order parameters for PE, PG and PS lipids in simulations}

@@

\subsection{PE headgroup}
Order parameters from Slipids simulations 
and experiments for DPPE are shown in Fig. \ref{HGorderParametersPE}.
Glycerol backbone order parameters in Slipids are off from experiments,
as already observed previously for PC lipids \cite{botan15}.
Order parameter signs for PE are not experimentally measured yet.
For headgroup the signs are set to give best agreement with simulations
and for glycerol to be consistent with experimental signs for PC.
Order parameter for $\beta$ carbon shows apparent agreement
with experiments. However, the sign of beta order parameter is positive,
in contrast to PC where negative sign was measured. Thus, the the beta order
parameter agrees with experiment with the assumption that its sign is opposite
than for PC. This is yet to be confirmed by experiments.
Order parameter for $\alpha$ carbon is too close to zero, even if the sign
would be correct.

@@

\subsection{PG headgroup}
Comparison between experiments and simulations for PG lipids is shown
in Fig. \ref{HGorderParametersPOPG}. The signs are not yet measured
experimentally. They are set to give the best argeement with experiments.
This would suggest that the $\beta$ order parameter would be positive,
in contrast to PC and PS headgroups, were negative signs were measured.
Even thought the signs turned out to be correct, the tested models would
not give a very good argeement with the experiments.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPG.eps}
  \caption{\label{HGorderParametersPOPG}
    Order parameters for PG headgroup and glycerol backbone from simulations and experiments without CaCl$_2$ 
    (POPG from \cite{borle85} contains 10mM of PIPES, DPPG from \cite{wohlgemuth80} contains 10mM PIPES and 100mM CaCl,
    E.Coli PG results from \cite{gally81}).
    Signs are not known for experimental data. They are determined to give best agreement
    with simulations. This is not reliable and should be corrected when experimental
    data becomes available.
  }
  \todo{More simulation data for lipids with different headgroups to be collected} \\
  \todo{CHARMM GUI simulation contains only counter ions as potassium.
    All experiments here contain some amount of sodium salt. The best
    ion concentrations for comparison should be figured out.} \\
  \todo{Experimental signs of the order parameters would highly useful.}
\end{figure}

\section{Lipid bilayers with mixed compositions}

\subsection{Effect of PE, PS and PG on PC headgroup}
The headgroup order parameters for PC lipids (POPC and DOPC)
mixed with PE, PS and PG lipids are shown in Fig. \ref{HGorderparametersPCvsPEPSPG}
from different simulation model and experiments \cite{scherer87} with different
mole fractions. As already discussed previosly, the PC lipid headgroup behaviour
follows the electrometer concept in experiments when mixed with other lipids, i.e., the order
parameters increase when mixed with negatively charged lipids (PS, PI, CL, PA and PG)
remains almost unchaged when mixed with neutral lipids (PE and SM) \cite{scherer87}.
This is not the case in simulation data shown in Fig. \ref{HGorderparametersPCvsPEPSPG}.
The addition of DOPE into a POPC and DOPC bilayers significantly decreases the PC headgroup
order parameters in simulations with OPLS compatible version of the Berger force field \cite{tieleman06}
in contrast to experiments \cite{scherer87}. On the other hand, the increase of the PC
headgroup order parameters in CHARMM36 simulations mixed with PS and PG lipids is significantly
smaller than in experiments.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
  \caption{\label{HGorderparametersPCvsPEPSPGchol}
    PC headgroup order parameters from experiments of mixtures with
    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
\end{figure}

@@

\section{Ca$^{2+}$ binding in bilayers with negatively charged PG and PS lipids}

PC lipid headgroup order parameters can used to measure ion binding
affinity, because their magnitude is linearly proportional
to the amount of bound charge in bilayer \cite{seelig87,catte16}.
This molecular electrometer concept can be used also
for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90}.
This is demonstrated in Figs \ref{OrderParametersWithCaCl},
\ref{OrderParametersWithCaClBELOW1M} and \ref{OrderParameterCHANGESWithCaClBELOW1M},
showing order parameters for PC headgroup $\alpha$ and $\beta$ carbons
as a function of CaCl$_2$ concentration in the presence of different amounts of
negatively charged PS or PG lipids.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaCl.eps}
  \caption{\label{OrderParametersWithCaCl}
    PC headgroup order parameters as a function of CaCl concentration from experiments containing charged lipids.
    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
    and POPC:POPG mixture data without NaCl from \cite{borle85}.
  }
  \todo{Check the NaCl concentrations in the samples.}
\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
  \caption{\label{OrderParametersWithCaClBELOW1M}
    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
  }
\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
    The change of PC headgroup order parameters
    in the presence of different amount of negatively charged lipids
    respect to the values without added CaCl$_2$.
    The original data is the same as in Fig. \ref{OrderParametersWithCaCl}. 
  }
\end{figure}

@@

\section{Ca$2+$ binding in bilayers with negatively charged PG and PS lipids in simulations}

Comparison of Ca$2+$ binding in PG between CHARMM36 simulations and experiments \cite{borle85}
is shown in Fig. \ref{changesWITHCaClPG}. The decrease of $\alpha$ order parameter
is in agreement with experiments, while decerase of $\beta$ order parameter is overestimated.
The result is very similar to the results with PC in NMRlipids II publication \cite{catte16}.
It should be, however, noted that the $\beta$-order parameters are not actually measured for PG,
but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
Anyway, the data presented in NMRlipids II project and in Fig. \ref{changesWITHCaClPG} together
suggest that Calcium binding is similarly overestimated by CHARMM36 model in pure POPC bilayers and mixtures with
POPG. The good agreement of $\alpha$ carbon would be explained by too weak dependence of its order
parameter of bound charge \todo{Should we check this against cationic surfactant experiments \cite{scherer89}?}.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaClPGPS.eps}
  \caption{\label{changesWITHCaClPG}
    PG order parameters as a function CaCl$_2$ concentration from experiments \cite{borle85} and CHARMM36 simulations.
    Note that beta order parameter is calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}, not actually measured. 
  }
\end{figure}

@@

@@

\section{Measurements of order parameter sign}

Fig. \ref{PShgSIGNS} summarizes the experimental results on the order parameter sign
measurement for POPS sample. The experimental protocol is the same used in Ref. \citenum{ferreira16}.
In (a) you see the headgroup region of the INEPT spectrum where alpha and beta are
identified. In (b) you have the R-PDLF slices for alpha and beta where you see one single
splitting for beta (which gives an order parameter equal to 0.12), and for alpha a superposition
of a large splitting (order parameter equal to 0.09) and a very small splitting which cannot be
calculated. On the bottom you have the S-DROSS slices of these two carbons. The grey lines show a
random collection of slices from noise such that it gets clear what is significant. The S-DROSS
slice for beta clearly shows that the order parameter is negative. The slice for alpha shows that
the higher order parameter is positive and suggests that the smaller order parameter is negative
(from the deviation towards negative values in the longer t1 times).
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PShgSIGNS.pdf}
  \caption{\label{PShgSIGNS}
    Experimental results for sign measurement for POPS sample
  }
\end{figure}

The results updated with SIMPSON simulations for the SDROSS profiles
are shown in Fig. \ref{PShgSIGNSsimpson}. The value for the smaller
alpha order parameter is taken from Fig 3 in Ref. \citenum{roux91},
because resolution in 13C NMR experiments was nor high enough to determine
numerical value for this. The plots in Fig. \ref{PShgSIGNSsimpson} (c) show
the following. The error bars and points are the experimental SDROSS data.
The thick lines are SIMPSON simulations. The simulations were done by using
the order parameter for beta equal to -0.12 and for alpha one order parameter
equal to 0.09 and the other equal to -0.02 (black) or 0.02 (grey).
Since the black lines agree with experimental data, we conlude that
the order parameters for $\beta$ carbon are -0.12 and for $\alpha$
order parameters are 0.09 and -0.02.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PShgSIGNSsimpson.pdf}
  \caption{\label{PShgSIGNSsimpson}
    Experimental results for sign measurement for POPS sample
  }
\end{figure}

\section{Dihedrals}
\begin{figure*}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/dihed1.png}
  \includegraphics[width=8.0cm]{../Figs/dihed2.png}
  \includegraphics[width=8.0cm]{../Figs/dihed3.png}
  \includegraphics[width=8.0cm]{../Figs/dihed4.png}
  \includegraphics[width=8.0cm]{../Figs/dihed5.png}
  \includegraphics[width=8.0cm]{../Figs/dihed6.png}
  \includegraphics[width=8.0cm]{../Figs/dihed7.png}
  \includegraphics[width=8.0cm]{../Figs/dihed8.png}
  \includegraphics[width=8.0cm]{../Figs/dihed9.png}
  \includegraphics[width=8.0cm]{../Figs/dihed10.png}
  \caption{\label{dihedrals}
    Experimental results for sign measurement for POPS sample
  }
\end{figure*}

>>>>

\title{NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids} %Title of paper

@@

In NMRlipids I and II project we were looking for a MD model
which would correctly reproduce headgroup and glycerol
backbone structures and cation binding for PC lipid bilayers \cite{botan15,catte16}.
Here we extend the same goal for other than PC lipids.
Currently the focus is on PE, PG and PS bilayers and their
mixtures with PC. Experimental data with different amounts of 
added salt is now collected and presented in this manuscript.

@@

@@

@@

\section{Results and Discussion}

\subsection{PE headgroup and glycerol backbone}
Order parameters from Slipids simulations 
and experiments for DPPE are shown in Fig. \ref{HGorderParametersPE}.
Glycerol backbone order parameters in Slipids are off from experiments,
as already observed previously for PC lipids \cite{botan15}.
Order parameter signs for PE are not experimentally measured yet.
For headgroup the signs are set to give best agreement with simulations
and for glycerol to be consistent with experimental signs for PC.
Order parameter for $\beta$ carbon shows apparent agreement
with experiments. However, the sign of beta order parameter is positive,
in contrast to PC where negative sign was measured. Thus, the the beta order
parameter agrees with experiment with the assumption that its sign is opposite
than for PC. This is yet to be confirmed by experiments.
Order parameter for $\alpha$ carbon is too close to zero, even if the sign
would be correct.

@@

\subsection{PG headgroup and glycerol backbone}
Comparison between experiments and simulations for PG lipids is shown
in Fig. \ref{HGorderParametersPOPG}. The signs are not yet measured
experimentally. They are set to give the best argeement with experiments.
This would suggest that the $\beta$ order parameter would be positive,
in contrast to PC and PS headgroups, were negative signs were measured.
Even thought the signs turned out to be correct, the tested models would
not give a very good argeement with the experiments.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPG.eps}
  \caption{\label{HGorderParametersPOPG}
    Order parameters for PG headgroup and glycerol backbone from simulations and experiments without CaCl$_2$ 
    (POPG from \cite{borle85} contains 10mM of PIPES, DPPG from \cite{wohlgemuth80} contains 10mM PIPES and 100mM CaCl,
    E.Coli PG results from \cite{gally81}).
    Signs are not known for experimental data. They are determined to give best agreement
    with simulations. This is not reliable and should be corrected when experimental
    data becomes available.
  }
  \todo{More simulation data for lipids with different headgroups to be collected} \\
  \todo{CHARMM GUI simulation contains only counter ions as potassium.
    All experiments here contain some amount of sodium salt. The best
    ion concentrations for comparison should be figured out.} \\
  \todo{Experimental signs of the order parameters would highly useful.}
\end{figure}

\subsection{Effect of PE, PS and PG on PC headgroup}
The headgroup order parameters for PC lipids (POPC and DOPC)
mixed with PE, PS and PG lipids are shown in Fig. \ref{HGorderparametersPCvsPEPSPG}
from different simulation model and experiments \cite{scherer87} with different
mole fractions. As already discussed previosly, the PC lipid headgroup behaviour
follows the electrometer concept in experiments when mixed with other lipids, i.e., the order
parameters increase when mixed with negatively charged lipids (PS, PI, CL, PA and PG)
remains almost unchaged when mixed with neutral lipids (PE and SM) \cite{scherer87}.
This is not the case in simulation data shown in Fig. \ref{HGorderparametersPCvsPEPSPG}.
The addition of DOPE into a POPC and DOPC bilayers significantly decreases the PC headgroup
order parameters in simulations with OPLS compatible version of the Berger force field \cite{tieleman06}
in contrast to experiments \cite{scherer87}. On the other hand, the increase of the PC
headgroup order parameters in CHARMM36 simulations mixed with PS and PG lipids is significantly
smaller than in experiments.
%\begin{figure}[]
%  \centering
%  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
%  \caption{\label{HGorderparametersPCvsPEPSPGchol}
%    PC headgroup order parameters from experiments of mixtures with
%    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
%    Signs are determined as discussed in \cite{botan15,ollila16}.
%  }
%\end{figure}

@@

\subsection{Ca$^{2+}$ binding in bilayers with negatively charged PG and PS lipids}

PC lipid headgroup order parameters can used to measure ion binding
affinity, because their magnitude is linearly proportional
to the amount of bound charge in bilayer \cite{seelig87,catte16}.
This molecular electrometer concept can be used also
for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90}.
This is demonstrated in Figs \ref{OrderParametersWithCaCl},
\ref{OrderParametersWithCaClBELOW1M} and \ref{OrderParameterCHANGESWithCaClBELOW1M},
showing order parameters for PC headgroup $\alpha$ and $\beta$ carbons
as a function of CaCl$_2$ concentration in the presence of different amounts of
negatively charged PS or PG lipids.
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaCl.eps}
%  \caption{\label{OrderParametersWithCaCl}
%    PC headgroup order parameters as a function of CaCl concentration from experiments containing charged lipids.
%    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
%    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
%    and POPC:POPG mixture data without NaCl from \cite{borle85}.
%  }
%  \todo{Check the NaCl concentrations in the samples.}
%\end{figure}
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
%  \caption{\label{OrderParametersWithCaClBELOW1M}
%    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
%  }
%\end{figure}
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
%  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
%    The change of PC headgroup order parameters
%    in the presence of different amount of negatively charged lipids
%    respect to the values without added CaCl$_2$.
%    The original data is the same as in Fig. \ref{OrderParametersWithCaCl}. 
%  }
%\end{figure}

@@

Comparison of Ca$2+$ binding in PG between CHARMM36 simulations and experiments \cite{borle85}
is shown in Fig. \ref{changesWITHCaClPG}. The decrease of $\alpha$ order parameter
is in agreement with experiments, while decerase of $\beta$ order parameter is overestimated.
The result is very similar to the results with PC in NMRlipids II publication \cite{catte16}.
It should be, however, noted that the $\beta$-order parameters are not actually measured for PG,
but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
Anyway, the data presented in NMRlipids II project and in Fig. \ref{changesWITHCaClPG} together
suggest that Calcium binding is similarly overestimated by CHARMM36 model in pure POPC bilayers and mixtures with
POPG. The good agreement of $\alpha$ carbon would be explained by too weak dependence of its order
parameter of bound charge
\todo{The response of CHARMM36 to cationic surfactant against experiments \cite{scherer89} to be checked.
  I have already ran the simulations, analysis to be done.}.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaClPGPS.eps}
  \caption{\label{changesWITHCaClPG}
    PG order parameters as a function CaCl$_2$ concentration from experiments \cite{borle85} and CHARMM36 simulations.
    Note that beta order parameter is calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}, not actually measured. 
  }
\end{figure}

@@

\section{Effect of Ca$2+$ binding on PG and PE headgroup}

@@

%\section{Measurements of order parameter sign}

%Fig. \ref{PShgSIGNS} summarizes the experimental results on the order parameter sign
%measurement for POPS sample. The experimental protocol is the same used in Ref. \citenum{ferreira16}.
%In (a) you see the headgroup region of the INEPT spectrum where alpha and beta are
%identified. In (b) you have the R-PDLF slices for alpha and beta where you see one single
%splitting for beta (which gives an order parameter equal to 0.12), and for alpha a superposition
%of a large splitting (order parameter equal to 0.09) and a very small splitting which cannot be
%calculated. On the bottom you have the S-DROSS slices of these two carbons. The grey lines show a
%random collection of slices from noise such that it gets clear what is significant. The S-DROSS
%slice for beta clearly shows that the order parameter is negative. The slice for alpha shows that
%the higher order parameter is positive and suggests that the smaller order parameter is negative
%(from the deviation towards negative values in the longer t1 times).
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/PShgSIGNS.pdf}
%  \caption{\label{PShgSIGNS}
%    Experimental results for sign measurement for POPS sample
%  }
%\end{figure}

%The results updated with SIMPSON simulations for the SDROSS profiles
%are shown in Fig. \ref{PShgSIGNSsimpson}. The value for the smaller
%alpha order parameter is taken from Fig 3 in Ref. \citenum{roux91},
%because resolution in 13C NMR experiments was nor high enough to determine
%numerical value for this. The plots in Fig. \ref{PShgSIGNSsimpson} (c) show
%the following. The error bars and points are the experimental SDROSS data.
%The thick lines are SIMPSON simulations. The simulations were done by using
%the order parameter for beta equal to -0.12 and for alpha one order parameter
%equal to 0.09 and the other equal to -0.02 (black) or 0.02 (grey).
%Since the black lines agree with experimental data, we conlude that
%the order parameters for $\beta$ carbon are -0.12 and for $\alpha$
%order parameters are 0.09 and -0.02.
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/PShgSIGNSsimpson.pdf}
%  \caption{\label{PShgSIGNSsimpson}
%    Experimental results for sign measurement for POPS sample
%  }
%\end{figure}

%\section{Dihedrals}
%\begin{figure*}[]
%  \centering
%  \includegraphics[width=8.0cm]{../Figs/dihed1.png}
%  \includegraphics[width=8.0cm]{../Figs/dihed2.png}
%  \includegraphics[width=8.0cm]{../Figs/dihed3.png}
%  \includegraphics[width=8.0cm]{../Figs/dihed4.png}
%  \includegraphics[width=8.0cm]{../Figs/dihed5.png}
%  \includegraphics[width=8.0cm]{../Figs/dihed6.png}
%  \includegraphics[width=8.0cm]{../Figs/dihed7.png}
%  \includegraphics[width=8.0cm]{../Figs/dihed8.png}
%  \includegraphics[width=8.0cm]{../Figs/dihed9.png}
%  \includegraphics[width=8.0cm]{../Figs/dihed10.png}
%  \caption{\label{dihedrals}
%    Experimental results for sign measurement for POPS sample
%  }
%\end{figure*}
